---
title: "Phenology Calibration with PhenoMeNals"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Phenology Calibration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


## 📦 These are the packages we will use....
```{r table, include=F, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, include=TRUE, warning=FALSE, message=FALSE}
library(phenomenals)
library(kableExtra)
library(ggplot2)
library(dplyr)
library(lubridate)
```

## 🌱 Goal

This vignette demonstrates how to calibrate phenological parameters for a grapevine variety and site using the `phenologyCalibration()` function.

We’ll use sample datasets included in the package:

- `colliOrientali`: daily weather data
- `bbchSample`: observed BBCH stages
- `phenomenalsParameters`: model parameters

## 📦 Load the data

```{r load-data}
data(colliOrientali)
data(bbchSample)
parameters <- phenomenalsParameters
```

## 🔍 Inspect the inputs
```{r inspect-inputs}
kable(head(colliOrientali), caption = "Example weather data") %>%
  kable_styling(font_size = 10)
kable(head(bbchSample), caption = "Example BBCH data") %>%
  kable_styling(font_size = 10)
```

## ⚙️ Run phenologyCalibration()
```{r run-calibration}
calib_result <- phenologyCalibration(
  weather_data = colliOrientali,
  referenceBBCH = bbchSample,
  phenomenalsParameters =phenomenalsParameters,
  start_year = 2006,
  end_year = 2012,
  sites = "ColliOrientali",
  varieties = "Carmenere",
  iterations = 100,
  timestep = "daily"
)
```

This will:

- Use the small demo weather and BBCH dataset  
- Calibrate parameters only for *Carmenere* at *ColliOrientali*  
- Run 100 iterations (feel free to increase for real applications, the higher the iterations, the longer the time to relax a bit)
- Takes around  

## 📤 Inspect output object

```{r result-structure}
str(calib_result, max.level = 1)
```

This helps users see:

- What’s inside (parameters, phenology)  
- That it’s a named list  

## 📈 Calibrated parameters

```{r view-parameters}
filtered_params <- calib_result$parameters |>
  dplyr::filter(site == "colliorientali", variety == "carmenere")

knitr::kable(filtered_params, caption = "Calibrated parameters for Carmenere at ColliOrientali")
```


All calibrated parameters are within their defined bounds. 🎯

However, if any parameter is very close to its minimum or maximum limit, it's good practice to slightly expand the calibration range and re-run the process. This helps avoid **boundary effects** and ensures the optimization isn’t artificially constrained.

> 📝 Tip: Parameters like `bbch65` or `bbch85` that consistently hit the edge of their range may indicate that your biological assumptions need refining, or that more calibration iterations are needed.

## 🍇 Phenology curve

This is where you show the time series of predicted BBCH stages over time.
```{r view-results, fig.width = 7, fig.height=3}
ggplot(calib_result$phenology |> 
         mutate(Date=as.Date(Date,format='%m/%d/%Y'),
                year=year(Date)),aes(x=Date)) + 
  geom_line(aes(y=BBCHPhase))+
  geom_point(aes(y=BBCHRef))+
  theme_classic()+
  geom_line(aes(y=ChillState),col='blue4')


```
  



The `phenologyCalibration()` function successfully fits thermal and stress-based parameters
to reproduce observed BBCH stages. This calibrated output can now be passed to
`runPhenomenals()` for trait modeling.
